---
title: "Thaum Pangenome Analysis: By Component"
output: html_notebook
---
## Preliminary
Load necessary packages
```{r}
library(crabrawlR)
library(stringr)
library(ggplot2)
library(reshape2)
library(micropan) # calculate pangenome size estimates


```


Load pangenome matrix and "clean" it (remove NAs), load taxa key
```{r}
# headers are incorrectly imported by read.csv, so set header=FALSE
# setwd("/Users/RoxanneB/Desktop/thaum-explorer")
pangenome = read.csv("panmatrices/Run8_pangenome_matrix_t0.tab", sep="\t", header=FALSE, stringsAsFactors=FALSE) 
pangenome = panpretty(pangenome)

# check and delete colummns with NAs due to incorrect importing. pangenome_clean has NA cols removed.
pangenome_clean = NAcheck(pangenome)

# read in taxa key
taxkey = read.table("taxalists/taxlistR8.tab", header=TRUE, stringsAsFactors=FALSE)

# read in COG annotation table
thaumtab = read.table("annotation/R8ThaumAnno.tab", sep="\t", quote = "", stringsAsFactors=FALSE)
rownames(thaumtab) = thaumtab[,1]

categories = c("D", "M", "N", "O", "T", "U", "V", "W", "X", "Y", "Z", "A", "B", "J", "K", "L", "C", "E", "F", "G", "H", "I", "P", "Q", "R", "S")

```

## Total counts of each pangenome component
Separate taxa by phylum status (Thaum vs. Aig).
```{r}
thaum = taxkey[which(taxkey$Type_Phylum == "thaum"),1]
aig = taxkey[which(taxkey$Type_Phylum == "aig"),1]

taxlistThaum = taxrownums(df=pangenome_clean, taxa_yes=thaum)
taxlistAig = taxrownums(df=pangenome_clean, taxa_yes=aig)

```

Calculate Shared and Unique Core/Soft-Core/Cloud/Shell of Thaum Pangenome
```{r}
sharedpan = sharedgroups(pangenome_clean, taxlistThaum$r_yes)
uniquepan = getunique(pangenome_clean, taxlistThaum$r_yes, taxlistAig$r_yes)
```



Summary dataframe of counts (for supplementary tables, etc ...)
```{r}
byElement = data.frame(Shared=c(sharedpan$yes_strict_ct, sharedpan$yes_soft_ct, sharedpan$yes_shell_ct, sharedpan$yes_cloud_ct, sharedpan$yes_atleastone_ct),
                  Unique=c(uniquepan$yes_strict_ct, uniquepan$yes_soft_ct, uniquepan$yes_shell_ct, uniquepan$yes_cloud_ct, uniquepan$yes_atleastone_ct))
                           
                        
rownames(byElement) = c("strict", "soft", "shell", "cloud", "pangenome")

```


## Functional category enrichment

Phylum-wide COG category counts
```{r}
# setwd("/Users/RoxanneB/Desktop/thaum-explorer")
# Create dataframe of counts
pancounts = data.frame(matrix(ncol=0,nrow=0), stringsAsFactors=FALSE)
for(i in 1:length(categories)) {
  result = c(as.character(categories[i]), sum(str_detect(thaumtab$V3, categories[i]), na.rm=TRUE))
  pancounts = rbind(pancounts, result, stringsAsFactors=FALSE)
}

pancounts = rbind(pancounts, c("NA", sum(is.na(thaumtab$V3))))
rownames(pancounts) = pancounts[,1]
colnames(pancounts) = c("cog_category", "pan_total")
pancounts$pan_total = as.numeric(as.character(pancounts$pan_total))
pancounts$cog_category = factor(pancounts$cog_category, levels=pancounts$cog_category)

# Create barplot of categories (x-axis) vs. number of OFs in that category (y-axis)
ggplot(pancounts, aes(x=cog_category, y=pan_total)) + geom_bar(position="dodge", stat="identity", fill="turquoise3") + labs(y="# of ortholog families", x="COG category") + geom_text(aes(label=pan_total), size=3, vjust=-1)

ggsave('rplots/cogcatvsnumber_R8.pdf', plot = last_plot(), device = NULL, path = NULL,
  scale = 1, units = "in", width= 20, height= 10, dpi = 300, limitsize = TRUE)

# Remove Poorly Characterized & plot
pancounts_noPC = pancounts[which(rownames(pancounts) != "R" & rownames(pancounts) != "S" & rownames(pancounts) != "NA"),]

ggplot(pancounts_noPC, aes(x=cog_category, y=pan_total)) + geom_bar(position="dodge", stat="identity", fill="turquoise3") + labs(y="# of ortholog families", x="COG category") + geom_text(aes(label=pan_total), size=3, vjust=-1)

ggsave('rplots/cogcatvsnumber_R8-noPC.pdf', plot = last_plot(), device = NULL, path = NULL,
  scale = 1, units = "in", width= 20, height= 10, dpi = 300, limitsize = TRUE)


```

Core vs. Accessory (Strict and Soft)
```{r}
# Strict Core and Strict Accessory
## Retrieve ortholog family names for subsetting
nStrictCore = gsub(".faa", "", names(sharedpan$yes_strict))
nStrictAcc = gsub(".faa", "", names(sharedpan$yes_atleastone))
nStrictAcc = nStrictAcc[!nStrictAcc %in% nStrictCore]

strictCoreCounts = getcounts(nStrictCore, categories, thaumtab)
strictAccCounts = getcounts(nStrictAcc, categories, thaumtab)

## Set up contingency tables for FET (as dataframe)
strictCoreDF = data.frame(core_in_cat= strictCoreCounts,
                 core_not_cat= length(nStrictCore) - strictCoreCounts,
                 acc_in_cat= strictAccCounts,
                 acc_not_cat= length(nStrictAcc) - strictAccCounts)

strictAccDF = data.frame(acc_in_cat= strictAccCounts,
                 acc_not_cat= length(nStrictAcc) - strictAccCounts,
                 core_in_cat= strictCoreCounts,
                 core_not_cat= length(nStrictCore) - strictCoreCounts)

## Fisher Exact Test and Bonferroni correction
strictCoreP = apply(strictCoreDF,1, function(x) fisher.test(matrix(x,nr=2), alternative="greater")$p.value)
strictAccP = apply(strictAccDF,1, function(x) fisher.test(matrix(x,nr=2), alternative="greater")$p.value)

strictP= cbind(strictCoreP, strictAccP)
strictP_corrected = sapply(strictP, p.adjust, method="bonferroni")
strictP_corrected = as.data.frame(matrix(strictP_corrected, nc=2), stringsAsFactors=FALSE) 
rownames(strictP_corrected) = c(categories, "NA")
colnames(strictP_corrected) = c("strictCoreP", "strictAccP")

# Soft Core and Soft Accessory
## Retrieve ortholog family names for subsetting
nSoftCore = gsub(".faa", "", names(sharedpan$yes_soft))
nSoftAcc = gsub(".faa", "", names(sharedpan$yes_atleastone))
nSoftAcc = nSoftAcc[!nSoftAcc %in% nSoftCore]

softCoreCounts = getcounts(nSoftCore, categories, thaumtab)
softAccCounts = getcounts(nSoftAcc, categories, thaumtab)

## Set up contingency tables for FET (as dataframe)
softCoreDF = data.frame(core_in_cat= softCoreCounts,
                 core_not_cat= length(nSoftCore) - softCoreCounts,
                 acc_in_cat= softAccCounts,
                 acc_not_cat= length(nSoftAcc) - softAccCounts)

softAccDF = data.frame(acc_in_cat= softAccCounts,
                 acc_not_cat= length(nSoftAcc) - softAccCounts,
                 core_in_cat= softCoreCounts,
                 core_not_cat= length(nSoftCore) - softCoreCounts)

## FET and Bonferroni correction
softCoreP = apply(softCoreDF,1, function(x) fisher.test(matrix(x,nr=2), alternative="greater")$p.value)
softAccP = apply(softAccDF,1, function(x) fisher.test(matrix(x,nr=2), alternative="greater")$p.value)

softP= cbind(softCoreP, softAccP)
rownames(softP) = c(categories, "NA")
softP_corrected = sapply(softP, p.adjust, method="bonferroni")
softP_corrected = as.data.frame(matrix(softP_corrected, nc=2), stringsAsFactors = FALSE)
rownames(softP_corrected) = c(categories, "NA")
colnames(softP_corrected) = c("softCoreP", "softAccP")
```


# Now combine strict and soft
## rearrange so it's easier to read!! -- strictCore and softCore next to each other, same with accessory
```{r}
pangenomepartsP = cbind(strictP_corrected, softP_corrected)
pangenomepartsP = cbind(pangenomepartsP[,1], pangenomepartsP[,3], pangenomepartsP[,2], pangenomepartsP[,4]) 
colnames(pangenomepartsP) =c("strict_core", "soft_core", "strict_acc", "soft_acc")
rownames(pangenomepartsP) = c(categories, "NA")

# Pick out significant categories (p <.05)
pangenomepartsP_sig = which(pangenomepartsP < .05, arr.ind=T)
pangenomepartsP_sig_names = as.data.frame(cbind(rownames(pangenomepartsP_sig), colnames(pangenomepartsP[,pangenomepartsP_sig[,2]])), stringsAsFactors = FALSE)
colnames(pangenomepartsP_sig_names) = c("category", "pan_part")


pangenomepartsP_sig_sorted = aggregate(pan_part ~ category, pangenomepartsP_sig_names, function(gf) 
      paste(unique(unlist(strsplit(gf, ", "))), collapse = ',')) 


```

Plot stacked barplot (core + accessory stacked for each COG category)
```{r}
## Arrange strict
strict_wide = data.frame(corefreq=strictCoreCounts / (strictCoreCounts + strictAccCounts),
                         accfreq=strictAccCounts / (strictCoreCounts + strictAccCounts))

rownames(strict_wide) = c(categories, "NA")
strict_wide = na.omit(strict_wide)

strict_wide = cbind(rownames(strict_wide), strict_wide)
colnames(strict_wide) = c("category", "core_freq", "acc_freq")

strict_long = melt(strict_wide, id= c("category"))
colnames(strict_long) = c("category", "component", "freq")

strict_long$category = factor(strict_long$category, levels=strict_long$category)
# strict_wide = cbind(strict_wide, rep_len("C",length(strict_wide$strictCore))) -- only in oldstyle plotting

ggplot(strict_long, aes(fill=component, y=freq, x=category)) + geom_bar(stat= "identity") + labs(y="% in category", x="category")
ggsave('rplots/coreaccessorystack_R8_strict.pdf', plot = last_plot(), device = NULL, path = NULL,
  scale = 1, units = "in", width= 20 , height= 10, dpi = 300, limitsize = TRUE)


## Arrange soft
soft_wide = data.frame(corefreq=softCoreCounts / (softCoreCounts + softAccCounts),
                         accfreq=softAccCounts / (softCoreCounts + softAccCounts))

rownames(soft_wide) = c(categories, "NA")
soft_wide = na.omit(soft_wide)

soft_wide = cbind(rownames(soft_wide), soft_wide)
colnames(soft_wide) = c("category", "core_freq", "acc_freq")

soft_long = melt(soft_wide, id= c("category"))
colnames(soft_long) = c("category", "component", "freq")

soft_long$category = factor(soft_long$category, levels=soft_long$category)
# soft_wide = cbind(soft_wide, rep_len("C",length(soft_wide$softCore))) -- only in oldstyle plotting

ggplot(soft_long, aes(fill=component, y=freq, x=category)) + geom_bar(stat= "identity") + labs(y="% in category", x="category")
ggsave('rplots/coreaccessorystack_R8_soft.pdf', plot = last_plot(), device = NULL, path = NULL,
  scale = 1, units = "in", width= 20 , height= 10, dpi = 300, limitsize = TRUE)

```




## Stacked barplot -- old way of plotting with combined strict and soft w/ facets
```{r}


## Combine strict and soft and transform to long -- old way
CA_wide = rbind(strict_wide, soft_wide)
CA_long = melt(CA_wide, id= c("category"))
colnames(CA_long) = c("category", "component", "counts")

ggplot(CA_long, aes(fill=component, y=counts, x=type)) + geom_bar(stat="identity", position="fill", width=0.9) + facet_grid( ~ category) + theme(panel.spacing = unit(0, "lines")) + labs(y="% in category", x="core definition type")

ggsave('coreaccessorystack.pdf', plot = last_plot(), device = NULL, path = NULL,
  scale = 1, units = "in", width= 20 , height= 10, dpi = 300, limitsize = TRUE)


```

Pangenome size estimates
```{r}
panThaumOnly = pangenome_clean[taxlistThaum$r_yes,]
panThaumChao = chao(panThaumOnly)
panThaumHeaps = heaps(panThaumOnly, n.perm = 500)
panThaumRF = rarefaction(panThaumOnly, n.perm = 500)
panThaumRFplot = plot(panThaumRF)
panThaumRFsum = summary(panThaumRF)
```